Exploring Hillary Clinton's Emails\thanks{STATS 601 final project.}

### 2016-04-20 paper/paper.tex
Starting edits.

To try and move beyond the limitations of spectral clustering and capture some of the content of the emails as well as structure, we next turn to latent Dirichlet allocation (LDA).  LDA is a popular form of topic modeling that has historically been effective for identifying clusters or "topics" in text corpuses\footnote{note on estimation}.

LDA requires the number of topics to be specified as an input to the model, therefore one of the preliminary steps for topic model estimation is the selection of an appropriate number of topics.  Since our ultimate goal was to find sensible topics that could be used to understand the email corpus we initially turned to several coherence measures to perform our model selection, in particular we considered the UMass and UCI measures\footnote{add bib here}.  Coherence measures were developed as a way to quantify whether topics could make sense to a human observer.  However, the coherence of our topics didn't appear to vary much assuming we used more than 15 topics and less than 50.  As a result we turned to manual methods for selecting the number of topics that appeared to perform the best.  We estimated LDA for 5,10,...,75 topics and selected the model that appeared to be the most sensible to a human observer.  This left us with a 30 topic model.

Using our resulting topic model we want to answer a series of questions concerning our resulting topics and the email corpus.  First, can we identify a sensible structure in our topics that captures the structure we see with spectral clustering?  Second, can we identify informative content topics that correspond to real world events?  Finally, can we say anything about the email senders based on the topics discussed?

@@

To answer the first question above, we have to turn to human intuition.  While the original problem was too massive for a human to take on, the LDA algorithm has produced the prescribed number of topics with the most common words for each topic.  This looks like:

@@

\begin{table}[h]
\centering
\scalebox{0.8}{
\begin{tabular}{cccccc}
\toprule
Israel & Elections & Libya & Afghanistan & Int. Dev. & Obama \\
\midrule
israel & democrat & libya & afghanistan & develop & presid \\
isra & republican & secur & pakistan & state & obama \\
peace & american & travel & afghan & support & said \\
palestinian & polit & libyan & militari & global & hous \\
netanyahu & parti & iraq & general & program & white \\
east & elect & embassi & karzai & effort & administr \\
negoti & obama & attack & offici & intern & polici \\
arab & percent & kill & war & work & aid \\
state & candid & march & forc & includ & advis \\
\bottomrule
\end{tabular}
}
\caption{Some interesting content topics}
\label{tab:content_topics}
\end{table}

@@

While our content topics do appear to be sensible, we also want to get a sense of whether our topics appear in correct location in time.  For email we have a time stamp so we can sort our topic proportions by time to get a sense of whether our topics line up with real world events that correspond to the human interpretation of our topics.  See figure \ref{fig:topic_time_plots} for a plot of the mean monthly topic proportion fo the subset of content topics discussed above.

Comparing our topic proportion plots to real world news events, we can see that our topics do a good job of picking up on the correct events.  In the Isreal topic the first peak of the topic proportions corresponds to the July 26th, 2010 joint Israeli and Romanian helicopter disaster, while the second peak corresponds to the September 9th, 2011 attack on the Israel embassy in Egypt.  Similarly, for the election topic each of the peaks corresponds to an election in the US.  Finally, the first peak of the Libya topic corresponds to the fall of Gaddafi's adminstration in October of 2011, while the second peak corresponds to the Benghazi attack on September 11th, 2012.  Again, while this shows only a subset of the topics we find that this pattern persists throughout our content topics\footnote{Comparable topic proportion plots are available on the github page for this project.}.

\begin{figure}[h]
\centering
\begin{subfigure}[Israel]{.30\linewidth}
    \includegraphics[width=\linewidth]{../images/time_plot6.pdf}
    \caption{Israel} \label{fig:t1}
\end{subfigure}
\begin{subfigure}[Election]{.30\linewidth}
    \includegraphics[width=\linewidth]{../images/time_plot11.pdf}
    \caption{Election} \label{fig:t2}
\end{subfigure}
\begin{subfigure}[Libya]{.30\linewidth}
    \includegraphics[width=\linewidth]{../images/time_plot18.pdf}
    \caption{Libya} \label{fig:t3}
\end{subfigure}
\begin{subfigure}[Afghanistan]{.30\linewidth}
    \includegraphics[width=\linewidth]{../images/time_plot27.pdf}
    \caption{Afghanistan} \label{fig:t4}
\end{subfigure}
\begin{subfigure}[Int. Dev.]{.30\linewidth}
    \includegraphics[width=\linewidth]{../images/time_plot23.pdf}
    \caption{Int. Dev.} \label{fig:t5}
\end{subfigure}
\begin{subfigure}[Obama]{.30\linewidth}
    \includegraphics[width=\linewidth]{../images/time_plot25.pdf}
    \caption{Obama} \label{fig:t6}
\end{subfigure}
\caption{Visualization of the average monthly topic proportion for a subset of content topics over time.}
\label{fig:topic_time_plots}
\end{figure}

@@

\begin{table}[h]
\centering
\label{topic_sources}
\scalebox{0.70}{
\begin{tabular}{cccccc}
\toprule
Israel & Elections & Libya & Afghanistan & Int. Dev. & Obama \\
\midrule
Sidney Blumenthal & SB & Huma Abedin & Judith McHale & JM & SB \\
(SB) & & (HA) & (JM) & & \\
\noalign{\vskip 5mm}
HA & Philippe Reines & Wendy Sherman & HA & Melanne Verveer & PR \\
& (PR) & (WS) & & (MV) & \\
\noalign{\vskip 5mm}
Jake Sullivan & JM & JS & MV & Anne-Marie & Cherly Mills \\
(JS) & & & & Slaughter (AMS) & (CM)\\
\noalign{\vskip 5mm}
AMS & CM & Monica Hanley & JS & CM & HA \\
& & (MH) & & & \\
\noalign{\vskip 5mm}
Hillary Clinton & HA & SB & Richard Verma & JS & HC \\
(HC) & & & (RV) & & \\
\bottomrule
\end{tabular}
}
\caption{Primary senders for interesting topics}
\label{tab:top_senders}
\end{table}

@@

\begin{table}[h]
\centering
%\scalebox{0.7}{
\begin{tabular}{ccc}
\toprule
Source & Success Rate & Testing Observations \\
\midrule
Hillary Clinton & 0.74 & 849 \\
Philippe Reines & 0.06 & 34 \\
Claire Coleman & 0.52 & 27 \\
Lauren Jiloty & 0.42 & 71 \\
Huma Abedin & 0.36 & 376 \\
Jake Sullivan & 0.23 & 410 \\
Sidney Blumenthal & 0.29 & 87 \\
Cherly Mills & 0.31 & 491 \\
Anne-Marie Slaughter & 0.36 & 42 \\
Monica Hanley & 0.09 & 53 \\
Judith McHale & 0.25 & 28 \\
Robert Russo & 0.00 & 10 \\
Richard Verma & 0.26 & 19 \\
Wendy Sherman & 0.08 & 12 \\
Melanne Verveer & 0.30 & 30 \\
Lona Valmoro & 0.34 & 41 \\
\bottomrule
\end{tabular}
%}
\caption{Multinomial logistic regression successes}
\label{tab:mlr_success}
\end{table}

To relate these results back to our earlier analysis we want to look at the coefficients for the multinomial logistic regression model as this should give us a sense of how strongly each sender is associated with the different topics.  See figure \ref{fig:coefficients} for a heatmap of the resulting coefficient matrix\footnote{Note that the coefficients are standardized and we threshold all standardized coefficients above or below 2 to make the heatmap easier to interpret.}.

>>>>

To try and move beyond the limitations of spectral clustering and capture some of the content of the emails as well as structure, we next turn to latent Dirichlet allocation (LDA). LDA is a popular form of topic modeling that has historically been effective for identifying clusters or ``topics'' in text corpuses.\footnote{note on estimation}

LDA requires the number of topics to be specified as an input to the model, therefore one of the preliminary steps for topic model estimation is the selection of an appropriate number of topics. Since our ultimate goal is to find sensible topics that can be used to understand the email corpus, we initially turned to several coherence measures to perform our model selection. In particular, we considered the UMass and UCI measures.\footnote{add bib here} Coherence measures allow us to quantify whether estimated topics make sense to a human observer. However, the coherence of our topics did not appear to vary much across a wide range of reasonable topic numbers, from roughly 15 to 50 topics used. As a result we manually examined the results from models using a range of topics and attempted to select the most parsimonious model, resulting in a 30 topic model.

% After estimating our final topic model we want to answer a series of questions concerning our resulting topics and the email corpus.  First, can we identify a sensible structure in our topics that captures the structure we see with spectral clustering?  Second, can we identify informative content topics that correspond to real world events?  Finally, can we say anything about the email senders based on the topics discussed?

@@

After choosing and estimating our topic model, the first task is to check whether we can actually find meaning in the resulting topics. If topics correspond to meaningful issues or structures then we can use these results to make useful observations about the data.

This ability to then perform ``human'' analysis on the results is one of the key features of LDA. Bringing human intuition to bear on a 28,000 email corpus is nearly impossible, but by condensing the data into a handful of topics LDA has allowed us to leverage our human understanding of semantics and meaning to analyze the data.

Analysis is most easily done by examining the top words from each topics to find patterns and themes. Table~XX shows the top 10 words for each of the 30 estimated topics.

@@

\begin{table}[h] \centering
\begin{tabular}{cccccc}
  \toprule
  Israel & Elections & Libya & Afghanistan & Int. Dev. & Obama \\
  \midrule
  israel & democrat & libya & afghanistan & develop & presid \\
  isra & republican & secur & pakistan & state & obama \\
  peace & american & travel & afghan & support & said \\
  palestinian & polit & libyan & militari & global & hous \\
  netanyahu & parti & iraq & general & program & white \\
  east & elect & embassi & karzai & effort & administr \\
  negoti & obama & attack & offici & intern & polici \\
  arab & percent & kill & war & work & aid \\
  state & candid & march & forc & includ & advis \\
  \bottomrule
\end{tabular}
\caption{Some interesting content topics}
\label{tab:content_topics}
\end{table}

@@

While our content topics do appear to be sensible, we also want to get a sense of whether our topics appear in correct location in time.  For email we have a time stamp so we can sort our topic proportions by time to get a sense of whether our topics line up with real world events that correspond to the human interpretation of our topics.  See Figure~\ref{fig:topic_time_plots} for a plot of the mean monthly topic proportion fo the subset of content topics discussed above.

Comparing our topic proportion plots to real world news events, we can see that our topics do a good job of picking up on the correct events.  In the Isreal topic the first peak of the topic proportions corresponds to the July 26th, 2010 joint Israeli and Romanian helicopter disaster, while the second peak corresponds to the September 9th, 2011 attack on the Israel embassy in Egypt.  Similarly, for the election topic each of the peaks corresponds to an election in the US.  Finally, the first peak of the Libya topic corresponds to the fall of Gaddafi's adminstration in October of 2011, while the second peak corresponds to the Benghazi attack on September 11th, 2012.  Again, while this shows only a subset of the topics we find that this pattern persists throughout our content topics.\footnote{Comparable topic proportion plots are available on the github page for this project.}

\begin{figure}[htb]
\centering
\begin{subfigure}[Israel]{.30\linewidth}
    \includegraphics[width=\linewidth]{../images/time_plot6.pdf}
    \caption{Israel} \label{fig:t1}
\end{subfigure}
\begin{subfigure}[Election]{.30\linewidth}
    \includegraphics[width=\linewidth]{../images/time_plot11.pdf}
    \caption{Election} \label{fig:t2}
\end{subfigure}
\begin{subfigure}[Libya]{.30\linewidth}
    \includegraphics[width=\linewidth]{../images/time_plot18.pdf}
    \caption{Libya} \label{fig:t3}
\end{subfigure}
\begin{subfigure}[Afghanistan]{.30\linewidth}
    \includegraphics[width=\linewidth]{../images/time_plot27.pdf}
    \caption{Afghanistan} \label{fig:t4}
\end{subfigure}
\begin{subfigure}[Int. Dev.]{.30\linewidth}
    \includegraphics[width=\linewidth]{../images/time_plot23.pdf}
    \caption{Int. Dev.} \label{fig:t5}
\end{subfigure}
\begin{subfigure}[Obama]{.30\linewidth}
    \includegraphics[width=\linewidth]{../images/time_plot25.pdf}
    \caption{Obama} \label{fig:t6}
\end{subfigure}
\caption{Visualization of the average monthly topic proportion for a subset of content topics over time.}
\label{fig:topic_time_plots}
\end{figure}

@@

\begin{table}[htb] \centering \scriptsize
\begin{tabular}{cccccc}
  \toprule
  Israel & Elections & Libya & Afghanistan & Int. Dev. & Obama \\
  \midrule
  Sidney Blumenthal & SB & Huma Abedin & Judith McHale & JM & SB \\
  (SB) & & (HA) & (JM) & & \\
  \noalign{\vskip 5mm}
  HA & Philippe Reines & Wendy Sherman & HA & Melanne Verveer & PR \\
  & (PR) & (WS) & & (MV) & \\
  \noalign{\vskip 5mm}
  Jake Sullivan & JM & JS & MV & Anne-Marie & Cherly Mills \\
  (JS) & & & & Slaughter (AMS) & (CM)\\
  \noalign{\vskip 5mm}
  AMS & CM & Monica Hanley & JS & CM & HA \\
  & & (MH) & & & \\
  \noalign{\vskip 5mm}
  Hillary Clinton & HA & SB & Richard Verma & JS & HC \\
  (HC) & & & (RV) & & \\
  \bottomrule
\end{tabular}
\caption{Primary senders for interesting topics}
\label{tab:top_senders}
\end{table}

@@

\begin{table}[htb] \centering
\begin{tabular}{ccc}
  \toprule
  Source & Success Rate & Testing Observations \\
  \midrule
  Hillary Clinton & 0.74 & 849 \\
  Philippe Reines & 0.06 & 34 \\
  Claire Coleman & 0.52 & 27 \\
  Lauren Jiloty & 0.42 & 71 \\
  Huma Abedin & 0.36 & 376 \\
  Jake Sullivan & 0.23 & 410 \\
  Sidney Blumenthal & 0.29 & 87 \\
  Cherly Mills & 0.31 & 491 \\
  Anne-Marie Slaughter & 0.36 & 42 \\
  Monica Hanley & 0.09 & 53 \\
  Judith McHale & 0.25 & 28 \\
  Robert Russo & 0.00 & 10 \\
  Richard Verma & 0.26 & 19 \\
  Wendy Sherman & 0.08 & 12 \\
  Melanne Verveer & 0.30 & 30 \\
  Lona Valmoro & 0.34 & 41 \\
  \bottomrule
\end{tabular}
\caption{Multinomial logistic regression successes}
\label{tab:mlr_success}
\end{table}

To relate these results back to our earlier analysis we want to look at the coefficients for the multinomial logistic regression model as this should give us a sense of how strongly each sender is associated with the different topics.  See Figure~\ref{fig:coefficients} for a heatmap of the resulting coefficient matrix.\footnote{Note that the coefficients are standardized and then thresholded at $\pm2$ for illustrative purposes.}
